### 2025-10-07 Proposals/2025-NSF-EntropicGeometry/NarrativeBody.tex
Edits before sending draft to Alejandro

A core insight that is guiding this enterprise is that geometric structures in both classical and quantum mechanics are equivalent to entropic structures. In classical mechanics, a single degree of freedom is a two dimensional symplectic manifold. In this case, the symplectic form $\omega$ is a volume form and therefore specifying $\omega$ is equivalent to specifying the Liouville measure $\mu(U) = \int_U \omega$. In classical statistical mechanics, the entropy of a statistical ensemble is given by $S(\rho) = - \int \rho \log \rho d\mu$ where $\rho$ is the probability density associated to the ensemble. For a uniform distribution with support $U$, the entropy reduces to $S(\rho_U) = \log \mu(U)$. Since the logarithm is an invertible function, one can recover the measure from the entropy. That is, specifying the entropy associated to all uniform distributions allows us to reconstruct the Liouville measure, and therefore the symplectic form. In the general case, Darboux coordinates allow, at least locally, to factorize the space into multiple independent degrees of freedom, where the volume is the product of the corresponding areas. The corresponding entropic property is that the entropy for the joint distribution is the sum of the marginals if the independent degrees of freedom are independently distributed. Therefore, if the distribution of all possible marginals are specified, the symplectic form can be recovered.

@@

While the state spaces of classical (i.e. symplectic manifolds) and quantum mechanics (i.e. projective Hilbert spaces) are rather different, the space of ensembles of both have a similar structure. They are both convex subsets of a topological vector space, equipped with a strictly concave entropy function. The core idea, then, is to use ensembles as a way to generalize the notion of physical states, and the entropy to characterize the geometry in a way that is independent of the particulars of each physical theory. In the past year we have done significant preliminary work that shows that not only this approach is possible, but is extremely productive. The proposed work is to turn those preliminary insights in an axiomatic specification that is fully physically motivated. If successful, such an approach will provide the following benefits:
\begin{enumerate}
	\item clarify exactly what physical premises guarantee the basic geometric structures of physical theories, allowing a more principled formulation of current physical theories as well as the search for new ones;
	\item provide a generalized structure for information theory, possibly allowing future generalization of results from both classical and quantum information theory
	\item provide useful constraints to more precisely formulate physical theories in the infinite dimensional case.
\end{enumerate}

\section{Proposed research program}

The goal is to find a very limited set of axioms to define the entropic structure of a physical theory, which can be justified from physical considerations, show that it imposes a geometric structure on a generic space of statistical ensembles and that it recovers the standard geometric structures of classical and quantum mechanics. Furthermore, the same structure should be connected to topological and measure theoretic structures used in physics.

@@

We want to stress that finding the correct starting point is not just a mathematical issue, but also a physics problem. Let us explore this problem through an example. The space $M_1([0,1])$ of probability measures defined over the $[0,1]$ interval is an infinite dimensional vector space. As Aliprantis and Border note: (add reference) ``Since there is more than one topology of interest on an infinite dimensional space, the choice of topology is a key modeling decision that can have economic as well as technical consequences.'' Unbeknowst to many physicists, this is true for physics as well. If we choose a weak topology, we are saying the convergence of expectation values of all continuous functions is enough to determine convergence of ensembles. While this may seem plausible at first glance, we can find a weakly convergent sequence for which the entropy does not converge. If the entropic structure of a physical theory is really equivalent to its geometric structure, not only the entropy should be continuous, but it is likely that the topology should be the one generated by the entropic structure. Ultimately, answering these questions means understanding both the physical and mathematical consequences of those decisions.

Our preliminary work convinced us that the notion of statistical ensemble (i.e. the collection of all outputs of a preparation procedure) seems to be particularly suitable for generalization for a number of both practical and conceptual considerations. In terms of the math, the structure of statistical ensembles of both classical and quantum mechanics can be made to look very similar. From the experimental side, statistical ensembles are what we produce and measure experimentally. On the conceptual side, the reproducibility requirement of scientific investigation means that physical laws are of the form ``whenever I prepare a system according to this procedure, I will find this range of outcomes.'' That is, physical laws are relationships between ensembles. Therefore, it makes sense to require that a physical theory must \emph{at least} posit a space of ensembles. Our preliminary work shows that there are three, possibly four, distinct set of requirements for an ensemble space $\Ens$, the set of ensemble given by a physical theory.

The first requirement is that of experimental verifiability of the theory. This has been thoroughly investigated in our previous work that established a precise correspondence between experimental verifiable statements (i.e. statement associated with an experimental test that terminates successfully in finite time if and only if the statement is true) and the open set of a ``natural'' topology, and between theoretical statements (i.e. statement associated with an experimental test without termination guarantees) and Borel sets. (add citation) Real and integer numbers, for example, are recovered from an idealized model representing references of a reference system. The key insight is that the set theoretic operations represent a two-valued logic system (i.e. truth or falsehood of statements) while the topology keeps track of a three-valued logic system (i.e. interior, exterior and boundary represent respectively successful termination, unsuccessful termination and non-termination). This work gives us a solid foundation on which to build, and guarantees that an ensemble space $\Ens$ is, at least, a $\mathsf{T}_0$ second countable topological space.

The second requirement is the ability to perform statistical mixtures. Given two ensembles $\ens[a], \ens[b] \in \Ens$, we can always choose a probability $p \in [0,1]$ and select $\ens[a]$ $p$ percent of the times and $\ens[b]$ $\bar{p} = 1-p$ percent of the times. This gives the ensemble space a convex structure as $p \ens[a] + \bar{p} \ens[b] \in \Ens$ for all $\ens[a], \ens[b] \in \Ens$ and $p \in [0,1]$. This is insight is already used in the literature, particularly in Generalized Probabilistic Theories (GPTs). The main innovation in this respect is that we do not limit ourselves to finite combinations and we do not simply allow all infinite combinations (which can be shown to include physically untenable cases). Rather, we let the topology do the work. That is, infinite convex combinations exist only if they are the topological limit of finite convex combinations. On physical grounds, the mixing operation must be continuous in the natural topology defined by experimental verifiability, much like addition and scalar multiplication are continuous in a topological vector space, leading to the idea of a topological convex space. While we haven't found such a structure already studied in the literature, and it would be extremely useful to us to see what can be generalized from topological vector spaces to topological convex spaces, this is a fairly well defined mathematical question at this point, and we believe that our expertise in the mathematical foundations of physical theories is better spent in continuing finding other well defined starting points.

The third requirement, central to the work, is characterizing the variability of elements within a statistical ensemble. Though we imagine ensembles to be a collection of instances, these instances are never formalized as they need to be left unspecified to leave the theory general. However, the variability of the elements within an ensemble is physically significant as it will influence, for example, the variability of statistical measurements. As we showed in previous work, entropy quantifies the variability of the elements within a distribution, and which gives us a physically meaningful starting point. The primary goal of the proposed work, then, is to find what axiomatic properties must this variability indicator have, and to understand what are their conceptual, physical and mathematical implications.

The hardest conceptual problem we found is that the entropy is defined differently in each space. In a classical discrete space, the entropy is $-\sum p_i \log p_i$; in classical statistical mechanics is $- \int \rho \log \rho d\mu$ where $\mu$ is the Liouville measure; in quantum mechanics is $-\tr(\rho \log \rho)$. How can we give a single definition for these very different expressions? Note that the expressions are different precisely because they are in terms of the specific representation of ensembles in the specific theory. Our approach is to simply characterize the entropy by its properties with respect to the topology and the convex structure. In our preliminary work, for example, the entropy $S : \Ens \to \mathbb{R}$ has currently the following properties:
\begin{itemize}
	\item Continuity
	\item Strict concavity: $S(p\ens[a] + \bar{p} \ens[b]) \geq p S(\ens[a]) + \bar{p} S(\ens[b])$ with the equality holding if and only if $\ens[a] = \ens[b]$
	\item Upper variability bound: there exists a universal function $I(p_1, p_2)$ (i.e. the same for all ensemble spaces) such that $S(p\ens[a] + \bar{p} \ens[b]) \leq I(p, \bar{p}) + p S(\ens[a]) + \bar{p} S(\ens[b])$; if the equality holds, $\ens[a]$ and $\ens[b]$ are orthogonal, noted $\ens[a] \ortho \ens[b]$
	\item Mixtures preserve orthogonality: $\ens[a] \ortho \ens[b]$ and $\ens[a] \ortho \ens[c]$ if and only if $\ens[a] \ortho p \ens[b] + \bar{p} \ens[c]$ for any $p \in (0,1)$
\end{itemize}
These properties are justified by the physics. For example, if we are performing a statistical mixture between two ensembles, the average variability cannot decrease, which justifies the concavity. These relatively simple properties allow us, for example, to show that $I(p,\bar{p})$, the maximum entropy increase during mixing, must be, up to a multiplicative constant, the Shannon entropy $ - \left(p \log p + \bar{p} \log \bar{p}\right)$. The proof is general, meaning that if we are right that those axioms must hold for any physical theory, then the Shannon entropy is of truly general applicability. Moreover, in the case of both classical and quantum mechanics, states can be decomposed into orthogonal components, which means the upper variability is valid with an equal, and can be used to calculate the entropy from that decomposition. That allows to recover the expression for the entropy in the three different cases.

Our preliminary work shows that an entropy satisfying those axioms forces the convex space to be cancellative, meaning that it embeds into a vector space. That is, if we pick an internal point $\ens \in \Ens$, we can define a vector space $V_{\ens}$ such that there exists an embedding $\iota : \Ens \hookrightarrow  V_{\ens}$ such that $\iota(p \ens[a] + \bar{p} \ens[b]) = p \iota(\ens[a]) + \bar{p} \iota(\ens[b])$ and $\iota(\ens) = 0$. If we pick two different internal points $\ens[a],\ens[b] \in \Ens$, we can show that there is an invertible affine map between $V_{\ens[a]}$ and $V_{\ens[b]}$. That is, the ensemble space should be more properly thought as embedding in a affine space. We can also show that it is ``directionally bounded,'' meaning its intersection with any affine line is finite interval. It is an open question whether the current axioms are enough to prove that the embedding is continuous in a topological vector space, and whether the ensemble space is a bounded subset.

We can also show that the entropy can be used to define geometric structures directly on the ensemble space. For example, we can define the mixing entropy as $MS(\ens[a], \ens[b]) = S\left(\frac{1}{2}\ens[a] + \frac{1}{2} \ens[b]\right) - \left(\frac{1}{2} S(\ens[a]) + \frac{1}{2} S(\ens[b])\right)$. If we express the entropy in bits, the mixing entropy satisfies the following:
\begin{enumerate}
	\item non-negativity: $MS(\ens[a], \ens[b]) \geq 0$
	\item identity of indiscernibles: $MS(\ens[a], \ens[b]) = 0 \iff \ens[a]=\ens[b]$
	\item unit boundedness: $MS(\ens[a], \ens[b]) \leq 1$
	\item maximality of orthogonals: $MS(\ens[a], \ens[b]) = 1 \iff \ens[a] \ortho \ens[b]$
	\item symmetry: $MS(\ens[a], \ens[b]) = MS(\ens[b], \ens[a])$
\end{enumerate}
It is almost a distance function, as it does not satisfy the triangle inequality, but it recovers the Jensen-Shannon divergence (JSD) in both classical and quantum mechanics. While JSD is the square of a distance function in both classical and quantum information theory, our current axioms do not guarantee $MS$ to have this property in general. It is an open question whether enough physical constraints exist for that to be the case. The $MS$ can be used to create open balls, and show that the topology must be Hausdorff. An interesting conjecture is whether a Hausdorff second countable topological convex space is metrizable, given that this is true for topological vector spaces. The current axioms do not guarantee the open balls to be convex. If we were to find sufficient physically motivated conditions that imply that the entropic open balls are convex and generate the topology, as we suspect it should be, the topology would be locally convex and therefore the entropic structure also may give us enough structure to define calculus.

If we assume we are in finite dimensions and that the entropy is twice differentiable with respect to the convex coefficients, the strict concavity of the entropy means that the Hessian is negative definite which allows us to define a metric tensor. Let $V_{\ens[a]}$ be the embedding vector space with $\ens[a] \in \Ens$ as origin. This will be a differentiable manifold. On a linear coordinate chart we can define $g_{\ens}( v, w ) = - \frac{\partial^2 S}{\partial \ens ^2}( v, w )$ where $v$ and $w$ are elements of the tangent space at $\ens \in \Ens \subset V_{\ens[a]}$. Since all Hessian are symmetric, $g_{\ens}$ will be bilinear, symmetric and positive definite: it is a metric tensor and $V_{\ens[a]}$ is an Hessian manifold. In classical and quantum mechanics it recovers the Fisher information metric. The ultimate goal would be to show that, in general case, the embedding is continuous into a locally convex topological vector space so that this construction can be extended to the infinite dimensional case. This would essentially show that the entropy provides, in general, the geometric structure of the state space of any physical theory.

The fourth structure, which is left for future work, is the generalization of the Lie brackets given by the Poisson brackets in the classical case and the commutators in the quantum case. These describe the relationship between the generators of transformations that leave the geometric structure (i.e. symplectic form and inner product respectively) unchanged, which means, as outlined before, is equivalent to the geometric structure. That is, it describes the group of deterministic and reversible motions, those that leave the entropy unchanged. Given the tight connection between the entropic structure and the Lie algebraic structure, it is unclear whether this will be recovered automatically once the entropic structure is correctly axiomatized, or whether it will require additional axioms. For this reason, we postpone the study of this aspect.

Having given an overview of the specifics of the problems, the proposed work consists of first finalizing and publishing some of the preliminary results, which include the equivalence of geometric and entropic structures in physical theories, the ability to recover the Shannon entropy in the generalized case and the three different entropies in the discrete/continuous/quantum cases. At the same time, explore whether the axiomatic can be expanded, without sacrificing potentially interesting physical cases, to recover potentially useful properties, such as the metrizability and local convexity of the topology generated by the entropy.

@@

Generalized ensemble spaces + geometry defined from entropy; full set of axioms

@@

Have: 1. Topological structure; 2. Convex structure; 3. Geometric structure.

Open problems: 1. Entropic open balls can generate topology? 2. Are all axioms independent?   …

@@

Mention economists evidently aware of topological issues more than physicists.

\section{Proposed research program}

\subsection{Intellectual merit}

\section{Summary of proposed research program}

@@

\noindent
\textbf{Relationship to other commitments of PI:} PI Aidala has been continuously involved in high-energy experimental QCD since 2001.  Her separate line of theoretical research on the mathematical foundations of physics is relatively new, with her first publication in 2018~\cite{aop-phys-blueprint}. She has successfully split her efforts between her different research lines for the past six years, with her work related to the Assumptions of Physics project producing 14 peer-reviewed publications, 1 more submitted paper, and a published technical book since 2018.  She is moreover in the process of reducing the number of personnel she supervises on QCD research in order to have more time for her theoretical work---four of her Ph.D.~students in QCD completed their degrees in the last 18 months, and two more are expected to graduate before the start of the proposed project.

\section{Personnel}
The proposed research will be performed by PI Christine Aidala, her collaborator Gabriele Carcassi, and Ph.D.~student Tobias Thrien, all part of the Physics Department at the University of Michigan.  See also the section on ``Qualifications of the PI and Research Team'' below.  \\
TODO: Add Alejandro assuming he agrees.

@@

\section{Broader impacts}
In addition to disseminating the goals and results of the project through traditional means such as peer-reviewed publications and presentations at workshops and conferences, the goals and results of the project will be disseminated to wider audiences by leveraging existing social media and other outreach infrastructure that has already been developed for the broader Assumptions of Physics project, in particular, Carcassi's two YouTube channels, as described below in the section on qualifications of the research team.  One of them is aimed at the level of individuals with upper-level undergraduate training in physics and/or mathematics, and the second, more technical channel is aimed at professional researchers, for which he creates videos presenting detailed technical results or describing open technical questions, leveraging the expertise of others to help advance research goals.  Graduate student Tobias Thrien originally discovered the AoP project through Carcassi's YouTube channel while a Bachelor's student in Germany several years ago.  He has since completed a Master's in theoretical cosmology in Germany and is currently a first-year Ph.D.~student at U-M, having come to Michigan specifically to work on the AoP project with the PI and Carcassi.  Regarding the impact of posting open technical problems, our first collaborative paper that resulted from someone reaching out to us after seeing a video, on classical mechanics as the high-entropy limit of quantum mechanics, is currently under review at a journal~\cite{aop-classicallimit}.  We were previously completely unaware of the researcher and his work at the University of Innsbruck, Austria. Carcassi additionally writes monthly blog-style essays related to the AoP project and advertised through social media platforms such as Facebook and LinkedIn.  Thus, the results of the proposed research will reach a much wider audience than through the publications and conference talks alone.  Further dissemination of the open-research approach may inspire researchers in completely different areas to adopt a similar approach for their own projects, with potential acceleration of research progress in particular within interdisciplinary areas, where not all necessary expertise may be available within a single research community.

TODO: Paragraph about impact of unifying disparate branches of physics, and eventual impact on physics education.\\

\noindent
\textbf{Assessment of broader impacts:}  The reach and impact of videos related to the proposed work will be assessed by the number of views they receive and by other measures of audience engagement such as comments posted or subsequent email contacts.  The impact of unifying disparate branches of physics will need to be assessed on a longer-term timescale than that of the grant, but early indicators could include talk invitations and paper citations. \\
TODO: Consider ways to assess engagement in more detail, per last year's reviews.

>>>>

A core insight guiding this enterprise is that geometric structures in both classical and quantum mechanics are equivalent to entropic structures. In classical mechanics, a single degree of freedom is a two-dimensional symplectic manifold. In this case, the symplectic form $\omega$ is a volume form and therefore specifying $\omega$ is equivalent to specifying the Liouville measure $\mu(U) = \int_U \omega$. In classical statistical mechanics, the entropy of a statistical ensemble is given by $S(\rho) = - \int \rho \log \rho d\mu$ where $\rho$ is the probability density associated to the ensemble. For a uniform distribution with support $U$, the entropy reduces to $S(\rho_U) = \log \mu(U)$. Since the logarithm is an invertible function, one can recover the measure from the entropy. That is, specifying the entropy associated to all uniform distributions allows us to reconstruct the Liouville measure, and therefore the symplectic form. In the general case, Darboux coordinates allow us, at least locally, to factorize the space into multiple independent degrees of freedom, where the volume is the product of the corresponding areas. The corresponding entropic property is that the entropy for the joint distribution is the sum of the marginals if the independent degrees of freedom are independently distributed. Therefore, if the distributions of all possible marginals are specified, the symplectic form can be recovered.

@@

While the state spaces of classical (i.e.~symplectic manifolds) and quantum mechanics (i.e.~projective Hilbert spaces) are rather different, the space of ensembles of both have a similar structure. They are both convex subsets of a topological vector space, equipped with a strictly concave entropy function. The core idea, then, is to use ensembles as a way to generalize the notion of physical states, and the entropy to characterize the geometry in a way that is independent of the particulars of each physical theory. In the past year, we have done significant preliminary work that shows that not only is this approach possible, but it is extremely productive. The proposed work is to turn those preliminary insights in an axiomatic specification that is fully physically motivated. If successful, such an approach will provide the following benefits:
\begin{enumerate}
	\item clarify exactly what physical premises guarantee the basic geometric structures of physical theories, allowing a more principled formulation of current physical theories as well as the search for new ones;
	\item provide a generalized structure for information theory, possibly allowing future generalization of results from both classical and quantum information theory
	\item provide useful constraints to more precisely formulate physical theories in the infinite-dimensional case.
\end{enumerate}

\section{Proposed research program}
\subsection{Intellectual merit}

The goal is to find a very limited set of axioms to define the entropic structure of a physical theory that can be justified from physical considerations, show that this set of axioms imposes a geometric structure on a generic space of statistical ensembles and that it recovers the standard geometric structures of classical and quantum mechanics. Furthermore, the same structure should be connected to topological and measure theoretic structures used in physics.

@@

We want to stress that finding the correct starting point is not just a mathematical issue, but also a physics problem. Let us explore this problem through an example. The space $M_1([0,1])$ of probability measures defined over the $[0,1]$ interval is an infinite-dimensional vector space. As Aliprantis and Border note in the context of their own field of economics [add reference],  ``Since there is more than one topology of interest on an infinite dimensional space, the choice of topology is a key modeling decision that can have economic as well as technical consequences.'' Unbeknowst to many physicists, this is true for physics as well. If we choose a weak topology, we are saying the convergence of expectation values of all continuous functions is enough to determine convergence of ensembles. While this may seem plausible at first glance, we can find a weakly convergent sequence for which the entropy does not converge. If the entropic structure of a physical theory is really equivalent to its geometric structure, not only should the entropy be continuous, but it is likely that the topology should be the one generated by the entropic structure. Ultimately, answering these questions means understanding both the physical and mathematical consequences of those decisions.

Our preliminary work convinced us that the notion of statistical ensemble (i.e.~the collection of all outputs of a preparation procedure) seems to be particularly suitable for generalization for a number of both practical and conceptual considerations. In terms of the math, the structure of statistical ensembles of both classical and quantum mechanics can be made to look very similar. From the experimental side, statistical ensembles are what we produce and measure experimentally. On the conceptual side, the reproducibility requirement of scientific investigation means that physical laws are of the form ``whenever I prepare a system according to this procedure, I will find this range of outcomes.'' That is, physical laws are relationships between ensembles. Therefore, it makes sense to require that any physical theory must \emph{at least} posit a space of ensembles. Our preliminary work shows that there are three, possibly four, distinct sets of requirements for an ensemble space $\Ens$, the set of ensembles given by a physical theory.  The work proposed here focuses on further development of the third.

The first requirement is that of experimental verifiability of the theory. This has been thoroughly investigated in our previous work that established a precise correspondence between experimental verifiable statements (i.e.~statements associated with an experimental test that terminates successfully in finite time if and only if the statement is true) and the open set of a ``natural'' topology, and between theoretical statements (i.e.~statement associated with an experimental test without termination guarantees) and Borel sets~\cite{aop-book,aop-topExpDisting}. Real and integer numbers, for example, are recovered from an idealized model representing references of a reference system. The key insight is that the set-theoretic operations represent a two-valued logic system (i.e.~truth or falsehood of statements) while the topology keeps track of a three-valued logic system (i.e.~interior, exterior and boundary represent respectively successful termination, unsuccessful termination and non-termination). This work gives us a solid foundation on which to build, and guarantees that an ensemble space $\Ens$ is, at least, a $\mathsf{T}_0$ second-countable topological space.

The second requirement is the ability to perform statistical mixtures. Given two ensembles $\ens[a], \ens[b] \in \Ens$, we can always choose a probability $p \in [0,1]$ and select $\ens[a]$ $p$ percent of the times and $\ens[b]$ $\bar{p} = 1-p$ percent of the times. This gives the ensemble space a convex structure as $p \ens[a] + \bar{p} \ens[b] \in \Ens$ for all $\ens[a], \ens[b] \in \Ens$ and $p \in [0,1]$. This insight is already used in the literature, particularly in Generalized Probabilistic Theories (GPTs) [add refs]. The main innovation in this respect is that we do not limit ourselves to finite combinations and we do not simply allow all infinite combinations (which can be shown to include physically untenable cases). Rather, we let the topology do the work. That is, infinite convex combinations exist only if they are the topological limit of finite convex combinations. On physical grounds, the mixing operation must be continuous in the natural topology defined by experimental verifiability, much like addition and scalar multiplication are continuous in a topological vector space, leading to the idea of a topological convex space. We haven't found such a structure already studied in the literature. While it would be extremely useful to us to see what can be generalized from topological vector spaces to topological convex spaces, we do not believe that domain knowledge in physics is specifically valuable here and leave this work to others, as we believe that our expertise in the mathematical foundations of physical theories is better spent in continuing to find other well defined starting points.

The third requirement, central to the proposed work, is characterizing the variability of elements within a statistical ensemble. Though we imagine ensembles to be a collection of instances, these instances are never formalized as they need to be left unspecified to leave the theory general. However, the variability of the elements within an ensemble is physically significant as it will influence, for example, the variability of statistical measurements. As we showed in previous work~\cite{Carcassi:2021}, entropy quantifies the variability of the elements within a distribution, which gives us a physically meaningful starting point. The primary goal of the proposed work, then, is to find what axiomatic properties this variability indicator must have, and to understand their conceptual, physical and mathematical implications.

The hardest conceptual problem we found is that the entropy is defined differently in each space. In a classical discrete space, the entropy is $-\sum p_i \log p_i$; in classical statistical mechanics it is $- \int \rho \log \rho d\mu$ where $\mu$ is the Liouville measure; in quantum mechanics it is $-\tr(\rho \log \rho)$. How can we give a single definition for these very different expressions? Note that the expressions are different precisely because they are in terms of the specific representation of ensembles in the specific theory. Our approach is to simply characterize the entropy by its properties with respect to the topology and the convex structure. In our preliminary work, for example, the axiom defining entropy $S : \Ens \to \mathbb{R}$ currently has the following properties:
\begin{itemize}
	\item Continuity
	\item Strict concavity: $S(p\ens[a] + \bar{p} \ens[b]) \geq p S(\ens[a]) + \bar{p} S(\ens[b])$ with the equality holding if and only if $\ens[a] = \ens[b]$
	\item Upper variability bound: there exists a universal function $I(p_1, p_2)$ (i.e.~the same for all ensemble spaces) such that $S(p\ens[a] + \bar{p} \ens[b]) \leq I(p, \bar{p}) + p S(\ens[a]) + \bar{p} S(\ens[b])$; if the equality holds, $\ens[a]$ and $\ens[b]$ are orthogonal, noted $\ens[a] \ortho \ens[b]$
	\item Mixtures preserve orthogonality: $\ens[a] \ortho \ens[b]$ and $\ens[a] \ortho \ens[c]$ if and only if $\ens[a] \ortho p \ens[b] + \bar{p} \ens[c]$ for any $p \in (0,1)$
\end{itemize}
These properties are justified by the physics. For example, if we are performing a statistical mixture between two ensembles, the average variability cannot decrease, which justifies the concavity. These relatively simple properties allow us, for example, to show that $I(p,\bar{p})$, the maximum entropy increase during mixing, must be, up to a multiplicative constant, the Shannon entropy~\cite{Shannon} $ - \left(p \log p + \bar{p} \log \bar{p}\right)$. The proof is general, meaning that if we are right that those axioms must hold for any physical theory, then the Shannon entropy is of truly general applicability. Moreover, in the case of both classical and quantum mechanics, states can be decomposed into orthogonal components, which means the upper variability is an equality and can be used to calculate the entropy from that decomposition. This allows us to recover the expressions for the entropy in the three different cases, classical discete, classical continuous, and quantum.

Our preliminary work shows that an entropy satisfying those properties forces the convex space to be cancellative, meaning that it embeds into a vector space. That is, if we pick an internal point $\ens \in \Ens$, we can define a vector space $V_{\ens}$ such that there exists an embedding $\iota : \Ens \hookrightarrow  V_{\ens}$ such that $\iota(p \ens[a] + \bar{p} \ens[b]) = p \iota(\ens[a]) + \bar{p} \iota(\ens[b])$ and $\iota(\ens) = 0$. If we pick two different internal points $\ens[a],\ens[b] \in \Ens$, we can show that there is an invertible affine map between $V_{\ens[a]}$ and $V_{\ens[b]}$. That is, the ensemble space should be more properly thought as embedding into an affine space. We can also show that it is ``directionally bounded,'' meaning its intersection with any affine line is a finite interval. It is an open question whether the currently defined properties are enough to prove that the embedding is continuous in a topological vector space, and whether the ensemble space is a bounded subset.

We can also show that the entropy can be used to define geometric structures directly on the ensemble space. For example, we can define the mixing entropy as $MS(\ens[a], \ens[b]) = S\left(\frac{1}{2}\ens[a] + \frac{1}{2} \ens[b]\right) - \left(\frac{1}{2} S(\ens[a]) + \frac{1}{2} S(\ens[b])\right)$. If we express the entropy in bits, the mixing entropy satisfies the following:
\begin{itemize}
	\item non-negativity: $MS(\ens[a], \ens[b]) \geq 0$
	\item identity of indiscernibles: $MS(\ens[a], \ens[b]) = 0 \iff \ens[a]=\ens[b]$
	\item unit boundedness: $MS(\ens[a], \ens[b]) \leq 1$
	\item maximality of orthogonals: $MS(\ens[a], \ens[b]) = 1 \iff \ens[a] \ortho \ens[b]$
	\item symmetry: $MS(\ens[a], \ens[b]) = MS(\ens[b], \ens[a])$
\end{itemize}
It is almost a distance function, as it does not satisfy the triangle inequality, but it recovers the Jensen-Shannon divergence (JSD) in both classical [add ref] and quantum [add ref] mechanics. While the JSD is the square of a distance function in both classical and quantum information theory, our current axioms do not guarantee $MS$ to have this property in general. It is an open question whether enough physical constraints exist for that to be the case. The $MS$ can be used to create open balls and show that the topology must be Hausdorff. An interesting conjecture is whether a Hausdorff second-countable topological convex space is metrizable, given that this is true for topological vector spaces. The current axioms do not guarantee the open balls to be convex. If we were to find sufficient physically motivated conditions that imply that the entropic open balls are convex and generate the topology, as we suspect should be the case, the topology would be locally convex and therefore the entropic structure also may give us enough structure to define calculus.

If we assume we are in finite dimensions and that the entropy is twice differentiable with respect to the convex coefficients, the strict concavity of the entropy means that the Hessian is negative definite, which allows us to define a metric tensor. Let $V_{\ens[a]}$ be the embedding vector space with $\ens[a] \in \Ens$ as origin. This will be a differentiable manifold. On a linear coordinate chart we can define $g_{\ens}( v, w ) = - \frac{\partial^2 S}{\partial \ens ^2}( v, w )$ where $v$ and $w$ are elements of the tangent space at $\ens \in \Ens \subset V_{\ens[a]}$. Since all Hessians are symmetric, $g_{\ens}$ will be bilinear, symmetric and positive definite: it is a metric tensor and $V_{\ens[a]}$ is a Hessian manifold. In classical and quantum mechanics it recovers the Fisher information metric. The ultimate goal would be to show that, in the general case, the embedding is continuous into a locally convex topological vector space so that this construction can be extended to the infinite-dimensional case. This would essentially show that the entropy provides, in general, the geometric structure of the state space of any physical theory.

The fourth requirement, which is left for future work, is the generalization of the Lie brackets for a generic ensemble space, given by the Poisson brackets in the classical case and the commutators in the quantum case. These describe the relationship between the generators of transformations that leave the geometric structure (i.e.~symplectic form and inner product respectively) unchanged, which means, as outlined before, leaving the entropic structure unchanged. That is, it describes the group of deterministic and reversible motions, those that leave the entropy unchanged~\cite{aop-HamConsInfoEnt,aop-HamQuantInfo,aop-book}. Given the tight connection between the entropic structure and the Lie-algebraic structure, it is unclear whether this will be recovered automatically once the entropic structure is correctly axiomatized, or whether it will require additional axioms. For this reason, we postpone the study of this aspect.

Having given an overview of the specifics of the problems, the proposed work consists of first finalizing and publishing some of the preliminary results, which include the equivalence of geometric and entropic structures in physical theories, the ability to recover the Shannon entropy in the generalized case and the three different entropies in the discrete/continuous/quantum cases. At the same time, we will explore whether the axiomatics can be expanded without sacrificing potentially interesting physical cases in order to recover potentially useful properties, such as the metrizability of the space and local convexity of the topology generated by the entropy.

@@

@@

@@

\section{Summary of proposed research program}
[Add brief recap]

@@

\noindent
\textbf{Relationship to other commitments of PI:} PI Aidala has been continuously involved in high-energy experimental QCD since 2001.  Her separate line of theoretical research on the mathematical foundations of physics is relatively new, with her first publication in 2018~\cite{aop-phys-blueprint}. She has successfully split her efforts between her different research lines for the past seven years, with her work related to the Assumptions of Physics project producing 15 peer-reviewed publications, 1 more submitted paper, and a published technical book since 2018.  She is moreover in the process of reducing the number of personnel she supervises on QCD research in order to have more time for her theoretical work---four of her Ph.D.~students in QCD completed their degrees in the last 18 months, and two more are expected to graduate before the start of the proposed project.

\section{Personnel}
The proposed research will be performed by PI Christine Aidala, her collaborator Gabriele Carcassi, and Ph.D.~student Tobias Thrien, all part of the Physics Department at the University of Michigan.  See also the section on ``Qualifications of the PI and Research Team'' below.  \\
TODO: Add Alejandro if he agrees.

@@

\subsection{Broader impacts}
In addition to disseminating the goals and results of the project through traditional means such as peer-reviewed publications and presentations at workshops and conferences, the goals and results of the project will be disseminated to wider audiences by leveraging existing social media and other outreach infrastructure that has already been developed for the broader Assumptions of Physics project, in particular, Carcassi's two YouTube channels, as described below in the section on qualifications of the research team.  One of them is aimed at the level of individuals with upper-level undergraduate training in physics and/or mathematics, and the second, more technical channel is aimed at professional researchers, for which he creates videos presenting detailed technical results or describing open technical questions, leveraging the expertise of others to help advance research goals.  Graduate student Tobias Thrien originally discovered the AoP project through Carcassi's YouTube channel while a Bachelor's student in Germany several years ago.  He has since completed a Master's in theoretical cosmology in Germany and is currently a second-year Ph.D.~student at U-M, having come to Michigan specifically to work on the AoP project with the PI and Carcassi.  Regarding the impact of posting open technical problems, our first collaborative paper that resulted from someone reaching out to us after seeing a video, on classical mechanics as the high-entropy limit of quantum mechanics, is currently under review at a journal~\cite{aop-classicallimit}.  We were previously completely unaware of the researcher and his work at the University of Innsbruck, Austria. Carcassi's August 2025 video on ``The need for physical mathematics'' has already received 49k views and led directly to an invitation to speak at the American Mathematical Society conference in Washington, DC in January 2026. Carcassi additionally writes monthly blog-style essays related to the AoP project and advertised through social media platforms such as Facebook and LinkedIn.  Thus, the results of the proposed research will reach a much wider audience than through the publications and conference talks alone.  Further dissemination of the open-research approach may inspire researchers in completely different areas to adopt a similar approach for their own projects, with potential acceleration of research progress in particular within interdisciplinary areas, where not all necessary expertise may be available within a single research community.

Carcassi and Aidala will host an annual online school that includes an introduction to generalized ensemble spaces and the mathematical relationships between geometric and entropic structures, showcasing various results from the proposed work in a didactic manner.  The school will consist of six hour-long lectures, two per day over three days, each followed by 30 minutes of questions and discussion, from 10 am - 1 pm U.S. Eastern Time, enabling individuals from across the Americas, Europe, Africa, and parts of Asia to participate at a reasonable hour.  The school will be advertised primarily through social media.  Registrants will be given a Zoom link for participation, and those who choose not to register will be able to follow a livestream on YouTube (on Carcassi's technical channel) and ask questions through the livestream chat.  There will be no fee to participate.  All lectures and discussion will also be recorded and available through YouTube later. With 27 registrants for the inaugural Assumptions of Physics online school in June 2024, this grew to 167 registrants in June 2025, mainly undergraduate and graduate students and recent graduates in physics and/or mathematics, but also a handful of professors and other senior researchers.  Among the registrants were individuals from India, Bangladesh, Pakistan, Ethiopia, Egypt, Sudan, Bolivia, Brazil, and Singapore, in addition to North America and Europe. During the period of proposed work, we expect the school to reach several hundred individuals across similar demographics.

Beyond the immediate research goals, the proposed work will serve to unify what are currently disparate branches of physics---classical and quantum, and aspects of statistical mechanics.  Developing a common framework will pave the way for eventual new approaches to physics education, with less compartmentalized teaching across the core advanced undergraduate courses.  Finding general mathematical structures that any physical theory must have will also help constrain the search for new physical theories.  \\

\noindent
\textbf{Assessment of broader impacts:}  The reach and impact of videos related to the proposed work will be assessed by the number of views they receive and by other measures of audience engagement such as comments posted or subsequent email contacts, as well as potential talk invitations or proposals for collaboration.  The reach and impact of the school will be assessed by the number of registrants and their demographics, as well as the subsequent views of the recorded lectures.  The impact of unifying disparate branches of physics will need to be assessed on a longer-term timescale than that of the grant, but early indicators could include talk invitations and paper citations. \\
